---
title: "data"
author: "Anh Pham"
date: "2023-11-28"
output:
  html_document:
    theme: united
    toc: yes
    toc_float: no
    number_sections: no
    highlight: tango
    fig_width: 10
    fig_height: 8
  word_document:
    toc: yes
  pdf_document:
    toc: yes
    fig_height: 6
    fig_width: 10
    number_sections: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load raw files from Kaggle dataset
```{r}
titanic_data <- read.csv("D:\\STAT 847\\Project\\Project\\Project\\train.csv", na.strings = c("", "NA", "NULL", " "))
```

# Data cleaning

Check the missing values by building bar chart
```{r}
library(ggplot2)

# Calculate the number of non-missing values per column
non_missing_counts <- colSums(!is.na(titanic_data))

# Convert to data frame for ggplot
non_missing_df <- data.frame(
  Feature = names(non_missing_counts),
  Count = non_missing_counts
)

# Plot using ggplot2
ggplot(non_missing_df, aes(x = Feature, y = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5, color = "black") +
  theme_minimal() +
  labs(title = "Non-Missing Values per Feature", x = "Feature", y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x labels if needed

```

Impute missing values in 'Age' and 'Embarked' columns:

```{r}
titanic_data$Age<-ifelse(is.na(titanic_data$Age), mean(titanic_data$Age, na.rm = TRUE), titanic_data$Age)

titanic_data$Embarked<-ifelse(is.na(titanic_data$Embarked), "S", titanic_data$Embarked)
```

Check missing values after imputation:

```{r}
library(ggplot2)

# Calculate the number of non-missing values per column
non_missing_counts <- colSums(!is.na(titanic_data))

# Convert to data frame for ggplot
non_missing_df <- data.frame(
  Feature = names(non_missing_counts),
  Count = non_missing_counts
)

# Plot using ggplot2
ggplot(non_missing_df, aes(x = Feature, y = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5, color = "black") +
  theme_minimal() +
  labs(title = "Non-Missing Values per Feature", x = "Feature", y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x labels if needed

```

```{r}
data <- titanic_data
```

## Data visualization

```{r}
# Create a contingency table for the number of survivors by gender
survival_gender_table <- table(Survived = data$Survived, data$Sex )

# Print the contingency table
print(survival_gender_table)
```

`Sex` vs `Survived` bar chart

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# First, convert it to a data frame
survival_gender_df <- as.data.frame.matrix(survival_gender_table)

# Add a 'Survived' column based on the row names
survival_gender_df$Survived <- as.numeric(row.names(survival_gender_df))
row.names(survival_gender_df) <- NULL

# Reshape the data to long format
long_data <- survival_gender_df %>%
  pivot_longer(cols = -Survived, names_to = "Gender", values_to = "Count")

# Now create the horizontal stacked bar chart
ggplot(long_data, aes(x = Gender, y = Count, fill = as.factor(Survived))) +
  geom_bar(stat = "identity", position = "stack") +
  coord_cartesian(ylim = c(0, 700)) +
  coord_flip() +  # Make the bars horizontal
  scale_fill_manual(values = c("0" = "#df6a74", "1" = "#53826a")) +
  labs(x = "Gender", y = "Number of Passengers", fill = "Survived") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

`Age` group vs `Survived` bar chart

```{r}
library(ggplot2)
library(dplyr)

data$AgeGroup <- cut(data$Age, breaks = seq(0, 100, by = 10), include.lowest = TRUE)

age_survival_counts <- data %>%
  group_by(AgeGroup, Survived) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Survival = ifelse(Survived == 1, "Survived", "Not Survived"))

# Plot the stacked area chart
ggplot(age_survival_counts, aes(x = AgeGroup, y = Count, fill = as.factor(Survival))) +
  geom_bar(stat='identity',position = 'stack') +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c("Not Survived" = "#df6a74", "Survived" = "#53826a")) +
  theme_minimal() +
  labs(x = "Age Group", y = "Number of Passengers", fill = "Survival Status") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if necessary

```

`Pclass` vs `Survived` bar chart
```{r}
library(ggplot2)
library(dplyr)


pclass_survival_counts <- data %>%
  group_by(Pclass, Survived) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Survival = ifelse(Survived == 1, "Survived", "Not Survived"))

# Plot the stacked area chart
ggplot(pclass_survival_counts, aes(x = Pclass, y = Count, fill = as.factor(Survival))) +
  geom_bar(stat='identity',position = position_dodge()) +
  geom_text(aes(label = Count), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, # Adjust this to position the label above the bar
            size = 3.5) +
  scale_fill_manual(values = c("Not Survived" = "#df6a74", "Survived" = "#53826a")) +
  theme_minimal() +
  labs(x = "Passenger class", y = "Number of Passengers", fill = "Survival Status") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))  # Rotate x-axis labels if necessary

```

`Embark` vs `Survived` bar chart
```{r}
library(ggplot2)
library(dplyr)

embark_survival_counts <- data %>%
  group_by(Embarked, Survived) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Survival = ifelse(Survived == 1, "Survived", "Not Survived"))
embark_survival_counts <- embark_survival_counts %>%
  mutate(Embarked = recode(Embarked, 'C' = 'Cherbourg', 'Q' = 'Queenstown', 'S' = 'Southampton'))
  
# Plot the stacked area chart
ggplot(embark_survival_counts, aes(x = Embarked, y = Count, fill = as.factor(Survival))) +
  geom_bar(stat='identity',position = 'stack') +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c("Not Survived" = "#df6a74", "Survived" = "#53826a")) +
  theme_minimal() +
  labs(x = "Point of Embarkation", y = "Number of Passengers", fill = "Survival Status") +
  theme(axis.text.x = element_text(angle = 45
                                   , hjust = 1))  # Rotate x-axis labels if necessary

```

`Sibsp` vs `Survived` bar chart

```{r}
library(ggplot2)
library(dplyr)

# Create a new column 'Has_Sibling' to indicate if the passenger has siblings/spouse aboard
data <- data %>%
  mutate(Has_Sibling = ifelse(SibSp > 0, "Yes", "No"))

# Now, calculate the counts for each combination of having a sibling or not and survival status
sibling_survival_counts <- data %>%
  group_by(Has_Sibling, Survived) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Survival = ifelse(Survived == 1, "Survived", "Not Survived"))
# Plot the stacked bar chart
ggplot(sibling_survival_counts, aes(x = Has_Sibling, y = Count, fill = Survival)) +
  geom_bar(stat = 'identity', position = position_dodge()) +
  geom_text(aes(label = Count), 
            position = position_dodge(width = 0.9), 
            hjust = -0.5, # Adjust this to position the label above the bar
            size = 3.5)+
  scale_fill_manual(values = c("Not Survived" = "#df6a74", "Survived" = "#53826a")) +
  coord_flip() +  # Make the bars horizontal
  theme_minimal() +
  labs(x = "Passenger with Siblings/ Spouse", y = "Number of Passengers", fill = "Survival Status") +
  ylim(0, 500) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1),legend.position = "bottom")

# Note: Adjust color values and theme settings as per your preference
```

`Parch` vs `Survived` bar chart

```{r}
library(ggplot2)
library(dplyr)

# Create a new column 'Has_Sibling' to indicate if the passenger has siblings/spouse aboard
data <- data %>%
  mutate(Has_Par = ifelse(Parch > 0, "Yes", "No"))

# Now, calculate the counts for each combination of having a sibling or not and survival status
par_survival_counts <- data %>%
  group_by(Has_Par, Survived) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(Survival = ifelse(Survived == 1, "Survived", "Not Survived"))

# Calculate proportions within each 'Has_Par' group for the bar plot
par_survival_counts <- par_survival_counts %>%
  group_by(Has_Par) %>%
  mutate(Proportion = Count / sum(Count))

# Create the bar plot for each 'Has_Par' group
ggplot(par_survival_counts, aes(x = Has_Par, y = Proportion, fill = Survival)) +
  geom_bar(stat = "identity", position = "dodge") +  # Use stat = "identity" and position = "dodge" for a bar plot
  geom_text(aes(label = Count), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, # Adjust this to position the label above the bar
            size = 3.5) +
  scale_fill_manual(values = c("Not Survived" = "#df6a74", "Survived" = "#53826a")) +
  labs(x = "Passenger with Parents/ Children", y = "Number of Passengers", fill = "Survival Status")+
  theme_minimal() +
  theme(legend.position = "bottom")
```

Correlation plot

```{r}
library(corrplot)
library(dplyr)
```

```{r}
data$PassengerId <- NULL
data$Name <- NULL
data$Cabin <- NULL
data$Ticket <- NULL
data$FamilySize <- NULL

numericVars <- select_if(data, is.numeric)
correlationMatrix <- cor(numericVars)
corrplot(correlationMatrix, method = "color", type = "lower", tl.col = "black", 
         diag = FALSE, addCoef.col = "black", number.cex = 0.7)
```
