---
title: "STAT-847 Group 2 Project Titanic Survival Prediction"
author: "Author: JAP"
date: "Date: 2023-11-18"
output: pdf_document
---

```{r message=FALSE, warning=FALSE, include=TRUE}
library(readxl)
library(ggplot2)
library(randomForest)
library(tree)
library(glmnet)
library(gridExtra)
library(nnet)
library(caret)
library(pROC)
library(klaR)
```



```{r , include=TRUE}
# # train data loading
library(titanic)
data("titanic_train")
MyTrainData<-titanic_train
colnames(MyTrainData)
MyTrainData<-MyTrainData[,-c(1,4,9, 11)]

#Adding value label for Survived variable
MyTrainData$Survived <- factor(MyTrainData$Survived, 
                              levels = c(0, 1),
                              labels = c("No", "Yes"))

#Adding value label for  the proxy for socio-economic status (SES) variable
MyTrainData$Pclass <- factor(MyTrainData$Pclass, 
                              levels = c(1, 2, 3),
                              labels = c("1st", "2nd", "3rd"))


#Re-coding sex variable in training data set
MyTrainData$Sex[MyTrainData$Sex == 'male'] <- 0
MyTrainData$Sex[MyTrainData$Sex == 'female'] <- 1
MyTrainData$Sex <- factor(MyTrainData$Sex, 
                              levels = c(0, 1),
                              labels = c("Male", "Female"))

#Re-coding Embarked variable in training data set
MyTrainData$Embarked[MyTrainData$Embarked == 'C'] <- 1
MyTrainData$Embarked[MyTrainData$Embarked == 'Q'] <- 2
MyTrainData$Embarked[MyTrainData$Embarked == 'S'] <- 3
MyTrainData$Embarked <- factor(MyTrainData$Embarked, 
                              levels = c(1, 2, 3),
                              labels = c("C", "Q", "S"))
```

***Missing value imputation :*** 
With respect to the imputation of missing values, we utilized a predictive modeling technique based on the distribution of the predictive variables. The predictive modeling approach was employed due to its ability to impute missing values by taking into account the interrelationships among variables in the dataset. Simple imputation techniques, such as replacing missing values with the mean or median, may introduce bias into the estimates for our specific context because the missing data were not eliminated arbitrarily (MCAR). For this reason, we believed that predictive modeling could mitigate this bias by incorporating alternative data-driven probabilities of values.


```{r , include=TRUE}
# Missing value imputation for the Embarked variables 
table (MyTrainData$Embarked , useNA = "ifany")
PredictData <- MyTrainData[is.na(MyTrainData$Embarked), ]
NonMissingData <- MyTrainData[!is.na(MyTrainData$Embarked), ]
multinom.model <- multinom(Embarked ~ Pclass + Sex + Age, data = NonMissingData)
PredictData$Embarked <- predict(multinom.model, newdata = PredictData, "class")
MyTrainData <- rbind(NonMissingData, PredictData)
table (MyTrainData$Embarked , useNA = "ifany")

# Missing value imputation for the Age variables of train data

names(MyTrainData)
dim(MyTrainData)
MyTrainData$Age<-ifelse(is.na(MyTrainData$Age), mean(MyTrainData$Age, na.rm = TRUE), MyTrainData$Age)
colSums(is.na(MyTrainData))


```


## Data partition

```{r include=TRUE}
set.seed(100)
trainIndex <- createDataPartition(MyTrainData$Survived, p = 0.8, list = FALSE, times = 1)
MyTrainDataSplit1 <- MyTrainData [trainIndex, ]
MyTrainDataSplit2 <- MyTrainData [-trainIndex, ]
```


## Naive Bayes Classification

Fitting naive bayes model, where Pclass, Sex, Age were used as independent variable

```{r include=TRUE}
library(e1071)
library(gmodels)
naive.model <- naiveBayes(Survived ~ ., data = MyTrainDataSplit1)
 naive.probs <- predict(naive.model, newdata = MyTrainDataSplit2, type = "raw")
 naive.probs1 <- predict(naive.model, newdata = MyTrainDataSplit2, type = "class")

```



```{r}
#Confusion Matrix for the fitted naive bayes model

confusion.matrix <- confusionMatrix(naive.probs1, MyTrainDataSplit2$Survived)
confusion.matrix

confusion.matrix.re2 <- table(naive.probs1, MyTrainDataSplit2$Survived)
confusion.matrix.re2

precision <- confusion.matrix.re2[1, 1] / sum(confusion.matrix.re2[1, ])
precision
recall <- confusion.matrix.re2[1, 1] / sum(confusion.matrix.re2[, 1])
recall
f1_score <- 2 * (precision * recall) / (precision + recall)
f1_score

#predicted_probs <- unlist(naive.probs)

#ROCurve<-roc(as.numeric(MyTrainDataSplit2$Survived),as.numeric(naive.probs))
#plot(ROCurve, main = "ROC Curve", col = "blue", lwd = 2)
#AUC <- round(auc(r), 3)
#AUC

#ROC_Object8 <- roc(MyTrainDataSplit2$Survived, naive.probs)
#AUC8 <- round(auc(ROC_Object8), 3)
```




